Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-01-05 analysis/paper_draft3.Rmd
update with Eric

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. Current methods for detecting bias make explicit, but unacknowledged, assumptions about how to incorporate historical strike data into the bias detection proess. Here, we show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how choice of alpha influences bias detection in the face of varying degrees of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application (R Shiny application) to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with real, historical strike data from a convenience sample of cases from one court.

@@

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is possible incompatibility between current and historical strike data for a given attorney that may arise because of missing data or because attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity). Current statistical approaches to the detection of bias have two extreme approaches to inclusion of historical data regardless of incompatibility between current and historical trials. Firstly,  historical data can be ignored entirely. This approach may lead to underpowered estimates of bias due to the low number of strikes in a single trial (typically 6, 10, or 15 strikes per party), and may lead to inflated estimates of bias. Secondly, historical and current trial data can be combined and analysed together. While this approach increases the sample size, it ignores the possible effects of incompatiblity between current and historical trial. Such incompatibility can lead to inflated type I error and reduced power [@Bennett2020].

@@

The use of the power prior in this context allows lawyers explicit control over the assumed degree of incompatibility between the current and historical trial data. Specifically, lower values (closer to 0) of the power prior parameter, $\alpha$, should be selected when incompatibility is expected to be high, and higher values (closer to 1) of the power prior parameter selected when incompatibility is low. Accordingly, the lawyer bringing the _Batson_ challenge chooses the value of $\alpha$ and must be able to explain and defend that choice as appropriately cooresponding to the likely degree of incompatibility between the current strike data and historical strike data available in the particular case.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. [editme] The goal of these methods is not to _prove_ bias, but rather to lend support to an inferrence of bias and a subsequent _Batson_ challenge. Previously used methods have either ignored historical trial data, or conducted analyses where current and historical data are merged, with no regard for possible incompatibility between trials. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can assess how sensitive that posterior estimate is to the available historical strike data. This is appealing because it makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether to bring a _Batson_ challenge or not.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias inference is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds illegal bias, because no approach to inferring bias based on historical strike data has that ability. In future work we hope to examine data-driven approached to the selection of $\alpha$; where the degree of incompatibility can be estimated and used to inform the choice of the degree of discounting to apply to historical data. [editme - refs?] Knowing how to quantify incompatibility and how to specify and appropriate prior $\alpha$ is an area of active research [e.g., https://arxiv.org/abs/2206.04379]

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal bias. Current methods for detecting bias make implicit, but unacknowledged, assumptions about how to incorporate historical strike data into the bias detection process. Here, we show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of alpha influences the estimation and detection of bias in the face of varying degrees of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application (R Shiny application) to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with real, historical strike data from a convenience sample of cases from one court.

@@

In turn, the estimation of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is possible incompatibility between current and historical strike data for a given attorney that may arise because of missing data or because attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity). Current statistical approaches to the detection of bias have two extreme approaches for inclusion of historical data. Firstly,  historical data can be ignored entirely. This approach may lead to underpowered estimates of bias due to the low number of strikes in a single trial (typically 6, 10, or 15 strikes per party), and may lead to inflated estimates of bias. Secondly, historical and current trial data can be combined and analysed together. While this approach increases the sample size, it ignores the possible effects of incompatiblity between current and historical trial. Such incompatibility can lead to inflated type I error and reduced power [@Bennett2020].

@@

The use of the power prior in this context allows lawyers explicit control over the assumed degree of incompatibility between the current and historical trial data.  A commonly adopted practice is to fix $\alpha$ and assess the sensitivity of results to different values [Carvalho & Ibrahim - add to bib]. Specifically, lower values (closer to 0) of the power prior parameter, $\alpha$, should be selected when incompatibility is expected to be high, and higher values (closer to 1) of the power prior parameter selected when incompatibility is low. Accordingly, the lawyer bringing the _Batson_ challenge chooses the value of $\alpha$ and must be able to explain and defend that choice as appropriately cooresponding to the likely degree of incompatibility between the current strike data and historical strike data available in the particular case. It should be noted that the purpose of this procedure is to estimate bias to support the bringing of a _Batson_ challenge. As such the senstivity of bias detection to the value of $\alpha$ is used to evaluate sensitivity of the estimation of bias and is not indicative of the sucess/failure of a _Batson_ challenge.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. [editme] The goal of these methods is not to _prove_ bias, but rather to lend support to an inferrence of bias and a subsequent _Batson_ challenge. Previously used methods have either ignored historical trial data, or conducted analyses where current and historical data are merged, with no regard for possible incompatibility between trials. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can assess how sensitive that posterior estimate is to the available historical strike data. This is appealing because it makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether or not to bring a _Batson_ challenge.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias estimation is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds illegal bias, because no approach to estimating bias based on historical strike data has that ability.

In future work we hope to examine data-driven approached to the selection of $\alpha$; where the degree of incompatibility can be estimated and used to inform the choice of the degree of discounting to apply to historical data. [editme - refs?] Knowing how to quantify incompatibility and how to specify and appropriate prior for  $\alpha$ is an area of active research [e.g., https://arxiv.org/abs/2206.04379, cite Eric]
